A Practical Optional Type System for Clojure

### 2012-10-24 dissertation/design-choices.tex
final dissertation (unreviewed by Rowan)

%Tobin-Hochstadt

% Problematic areas
% - flattened paired arguments. assoc, hash-maps, array-map
% - complex argument dependencies. comp, partial

@@

\subsection{Primitive Arrays}

% Primitive arrays are covariant, which is well-known to be statically unsound.
% We cannot trust the type signature of any Java method or field.
% Separate read and write types into two parameters.

@@

Protocols and datatypes are two of Clojure's main means of abstraction.

%section \ref{sec:datatypes}

@@

Typed Clojure takes a

Similar to Typed Racket, type inference is based on Local Type Inference
by Pierce and Turner~\cite{PT00}.

@@

Typed Clojure includes support for F-bounds on type variables\cite{CCHOM89}, as an extension
to Local Type Inference.

An extra environment from type variables to bounds is kept until after inference,
after which each unground bound is substituted with the inferred types and the
subtyping relationships are checked.

@@

Heterogeneous maps in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys.
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

They do not support concatenation (merging), as heterogeneous maps
only hold positive information on the keys with entries, not
which keys are \emph{absent}.
This makes concatenation unsafe, as keys to the right of a
concatentation can ``overwrite'' keys to the left. With no information
on the absence of keys, concatenation becomes an unsound operation.

>>>>

Strickland, Tobin-Hochstadt, and Felleisen's describe variable-arity polymorphism~\cite{STF09},
which has been implemented in Typed Racket.
Typed Clojure directly ports most of Typed Racket implementation of variable-arity polymorphism,
as described in section \ref{sec:impvariablearity}

Some common idiomatic Clojure functions like \lstinline|assoc|, \lstinline|hash-map|,
and \lstinline|partial| cannot be typed by Strickland et al./ .
Instead, Typed Clojure hard-codes some of these functions as primitives that cannot
be used in a higher-order context.
Expressing the types of these functions is set as future work (section \ref{sec:futurearity}).

@@

\subsection{Primitive Arrays}
\label{designarrays}

Primitive arrays in Java are covariant, which is well-known to be statically unsound~\cite{OW97,BK99}.
This mean that at compile time we cannot completely trust the type signature of any Java method or field that involves 
an array.

We handle this by introducing a type \lstinline|(Array c t)|, which is an array has the Java component type \lstinline|c| 
array that can read and write values of type \lstinline|t|.
Internally, \lstinline|Array| types expand to \lstinline|(Array3 c w r)|,
which is an array has the Java component type \lstinline|c|,
that can write type \lstinline|w| (which is contravariant in \lstinline|Array|) and read type \lstinline|r| 
(which is covariant in \lstinline|Array|).
\lstinline|(Array c t)| expands to \lstinline|(Array3 c t t)|.

\lstinline|Array3| is useful when an array is passed to/from Java or some untyped Clojure code.
As soon as an array leaves the safety of Typed Clojure, it can be considered to have type
\lstinline|(Array3 c Nothing t)|. It is illegal to write to an array of this type in Typed Clojure
without first checking its component type, but reads remain legal.

This scheme should be considered experimental, and is not completely fleshed out or implemented.
Notably, the component type \lstinline|c| seems redundant, and Typed Clojure currently
does not provide syntax to ``cast'' an unwriteable array to a be writeable.

@@

Protocols and datatypes are two of Clojure's main means of abstraction.
Protocols, have a set of unimplemented methods, are typed by
giving the type for each method.
Datatype definitions are type checked by ensuring the methods it
implements match the types defined by the Java interface or protocol
it is extending.

Common usages of datatypes are checkable, like datatype
constructors, field accesses, and protocol method invocation.
These are described in section \ref{sec:datatypes}.
Mutable fields in datatypes are not yet supported, but is
also a rarely used feature.

@@

Type inference in Typed Clojure is based on Local Type Inference
by Pierce and Turner~\cite{PT00}.
The main extension to Local Type Inference in Typed Clojure
is support for \emph{dotted variables} for variable-arity polymorphism.
This implementation is lifted directly from Typed Racket.

@@

Typed Clojure includes support for F-bounds on type variables\cite{CCHOM89}, as an extension
to Local Type Inference. 
Type variables can have upper and lower bounds, and F-bounds allow type variable bounds
to recursively refer to the type variable being bounded.
This is similar to Java and Scala, which both support F-bounded polymorphism.

@@

Heterogeneous maps types in Typed Clojure are simple enough to avoid
the complications associated with concatenable records (section \ref{sec:recordtypes})
while still supporting many Clojure idioms.
They support adding new entries with keyword keys, removing entries,
and lookup of keyword keys.
When using these operations with non-keyword keys, heterogeneous maps
are treated like plain hash-maps.

Heterogeneous map types do not support concatenation (merging), as they
only hold positive information on the keys with entries, not
which keys are \emph{absent}.
This makes concatenation unsafe, as keys to the right of a
concatenation can ``overwrite'' keys to the left. With no information
on the absence of keys, concatenation becomes an unsound operation.

### 2012-10-24 dissertation/experiments.tex
final dissertation (unreviewed by Rowan)

>>>>

Conduits are an advanced form of ``pipes'' using arrows, and a generalisation of monads~\cite{ConduitRef}.
\emph{Conduit} is a Clojure library developed by Duey and 
contributors~\footnote{Github home of Conduit: \url{https://github.com/jduey/conduit}} 
that supports programming with conduits.

The library had to be modified and simplified considerably to be able to type check.
Many existing conduits had troublesome variable-parameters, reminiscent of the
issues of assigning types to \lstinline|assoc| and \lstinline|partial|
(which is future work, see section \label{sec:futurearity}).
In these cases, simplified versions of the conduits were created that 
took fixed arguments.

An interesting property of the types assigned to conduits (which
are internally functions of a single argument) is that it is
a recursive type. This is not in itself interesting, but Typed Clojure
implements conduit types as \emph{type functions} with variance,
which required the variance of the conduit types to be known in advance.

Listing \ref{lst:conduittypes} shows the type constructor for conduit,
\lstinline|==>|. Notice the body of \lstinline|==>| includes a reference
to itself in a position where variance must be known in advance.
\lstinline|declare-alias-kind| is used to declare the kind of an alias.
Once an alias with an already declared kind is defined with \lstinline|def-alias|,
the defined kind must match the declared kind, otherwise a type error is thrown.
(The specifics of the types in listing \ref{lst:conduittypes} are not relevant to the rest of this discussion,
so they are not explained).

\begin{lstlisting}[caption=Types for Conduits, label=lst:conduittypes]
...

(def-alias Result
  (TFn [[x :variance :covariant]]
    (U nil ;stream is closed
       '[] ;abort/skip
       '[x];consume/continue
       )))

(def-alias Cont
  (TFn [[in :variance :covariant]
        [out :variance :invariant]]
    [(U nil [(Result in) -> (Result out)]) -> (Result out)]))

(declare-alias-kind ==> (TFn [[in :variance :contravariant]
                              [out :variance :invariant]] Any))

(def-alias ==>
  (TFn [[in :variance :contravariant] 
        [out :variance :invariant]]
    [in -> '[(U nil (==> in out)) (Cont out out)]]))

...
\end{lstlisting}

The resulting port of \emph{Conduit} was unsatisfying for real world use.
The simplifications made in order to type check the library reversed
any effort \emph{Conduit} made to conform to common Clojure idioms.
The future work on improving variable-argument types should reveal whether
this library can be ported satisfactorily.

### 2012-10-24 dissertation/lit-review.tex
final dissertation (unreviewed by Rowan)

\section{Java Interoperability}

Scala and Clojure are both languages with a focus on host interoperability, specifically
to the Java Virtual Machine. Their differing treatment of the Java Virtual Machine's \lstinline|null| is
relevant, particularly considering the potential of using occurrence typing~\cite{TF10}
to prevent erroneous usage of \lstinline|null|.

@@

Pierce and Turner split \emph{local type argument synthesis} into two further
algorithms: bounded, and unbounded quantification~\cite{PT00}. 
Typed Racket 
supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Piece et al.\ .
Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},
basing its type argument synthesis on the bounded quantification algorithm.

>>>>

\section{Java Interoperability in Statically Typed Languages}

Typed Clojure features integration with Java's static type system
to help verify interoperability with Java as correct.
Section \ref{sec:interactionnull} presents Typed Clojure's treatment of
Java's \emph{null} pointer. 
Section \ref{designarrays} shows how Typed Clojure deals with shortcomings
of Java's static type system relating to primitive arrays.

Other statically typed languages support Java interoperability with
similar goals, most completely, Scala.
Scala~\cite{OCD+} is tightly integrated with Java. It manages interactions
with Java's \emph{null} pointer by using an \emph{Option} type, which (if
used correctly) provides strong elimination for \emph{null}.
\emph{null} is still not expressible as a type in Scala, however.
Arrays are parameterised by a non-variant (or invariant) parameter,
which are checked using \emph{conservative approximation}~\cite{OSV08}. For example,
this checks that covariant parameters are only used in covariant positions.

@@

Pierce and Turner split \emph{local type argument synthesis} into two further
algorithms: bounded, and unbounded quantification~\cite{PT00}. 
Typed Racket 
supports unbounded polymorphism~\cite{Tob10}, implementing the latter algorithm by Pierce and Turner.
Scala supports bounded quantification with F-bounded polymorphism~\cite{CCHOM89},
basing its type argument synthesis on the bounded quantification algorithm.

### 2012-10-24 dissertation/paper.tex
final dissertation (unreviewed by Rowan)

Whether a language is dynamically typed or statically typed is starting to
blur. Recent languages have managed to support the advantages of both styles:
the safety of a statically typed language while retaining the idioms found in
dynamically typed languages.

Extensions like Typed Clojure

\begin{verbatim}
  Programming is an error-prone task, and verification tools are important.
  Dynamic languages rely on several techniques for verification, but static
  type checking is not often supported.
\end{verbatim}

>>>>

Whether a language is dynamically typed or statically typed is not always a
dividing classification. Recent languages have managed to support the advantages of both styles:
the safety of a statically typed language while retaining the idioms found in
dynamically typed languages.

This dissertation describes an optional static typing system
that attempts to bring the advantages of static typing to the dynamically type language
Clojure, running on the Java Virtual Machine. 
By reusing many of the design choices made in similar projects
like Typed Racket, we are able to design and implement a prototype type system
\emph{Typed Clojure} that can statically check many Clojure idioms.

Typed Clojure is intended to be of practical use to Clojure programmers.
We show that Typed Clojure help verify Clojure code written in sophisticated
programming styles is correct by porting most of a wide-spread Clojure library for monadic programming.
Similarly, we show that Typed Clojure helps verify code using Java interoperability is correct, and is
a useful tool when code involves usages of Java's \emph{null}.

There is still significant future work in order to type check all Clojure idioms,
but the work already carried out suggests that an optional type system for Clojure
like Typed Clojure is both practical and useful tool.
